---
title: "Palmer Penguins Initial Analysis"
author: "Darvin Ojha"
format: html
editor: visual
execute:
  keep-md: true
---

## PalmerPenguinsAnalysis

This is an analysis of the Palmer's Penguin data set. We begin with loading packages and datasets. We then ask interesting questions about the penguins throught the data set that was provided. These questions allow us to answer them in a unique and interacting way through different tools such as histograms, data chart, bargraph, etc.

## Loading Packages and Dataset

Here we will load the tidyverse package and penguins data.

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

#Read the penguins_samp1 data file from github
penguins <- read_csv("https://raw.githubusercontent.com/mcduryea/Intro-to-Bioinformatics/main/data/penguins_samp1.csv")

#See the first six rows of the data we've read in to our notebook
penguins %>% 
  head(2) %>% 
  kable() %>% 
  kable_styling(c("striped","hover"))
```

## About Our Data

The data we are working with is a dataset on Penguins, which includes 8 features measured on 44 penguins. The features included are physiological features (like bill length, bill depth, flipper length, body mass, etc) as well as other features like the year that the penguin was observed, the island the penguin was observed, the island the penguin was observed on, the sex of the penguin, and the species of the penguin.

## Interesting Questions to Ask

-   What is the average flipper length? What about for each species?

-   Are there more male or female penguins? What about per island or species?

-   What is the average body mass? What about by island? By species? By sex?

-   What is the ratio of bill length to bill depth for a penguin? What is the overall average of this metric? Does it change by species, sex, or island?

-   Does average body mass change by year?

    ## Data Manipulation Tools sand Strategies

We can look at individual columns in a data set or subsets of a column in a dataset. For example, if we are only interested in flipper length and species, we can `select()` those columns.

```{r}
penguins %>%
  select(species, body_mass_g)
```

If we want to `filter()` and only show certain rows, we can do that too.

```{r}
#We can filter by sex (catergorical variables)
penguins %>%
  filter(species =="Chinstrap")

#We can also filter by numerical varialbles
penguins %>%
  filter(body_mass_g >= 6000)

#We can filter 
penguins %>%
  filter((body_mass_g >=6000) | (island== "Torgersen"))
```

Using the filter(), we were able to filter out the data by sex (categorical variable). We also filtered the mass (numerical variable). We also filtered by island.

### **Answering Our Questions**

Most of our questions involve summarizing data, and perhaps summarizing over groups. We can summarize data using the `summarize()` function, and group data using `group_by()`. Here we asked several questions, perhaps what a biologist would ask in order to find out more about a certain species of organisms.

### Let's find the average flipper length.

```{r}
#Overall average flipper
penguins %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))
#Single Species Average 
penguins %>%
  filter(species == "Gentoo") %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

#Grouped Average
penguins %>%
  group_by(species) %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))
```

Using summarize(), we were able to calculate the average mean of the flipper length of the penguins. By using filter and summarize, we calculated the average flipper length of just the Gentoo species of Penguins. Finally using group, we categorized the average flipper length of all three species in a table.

### How many of each species do we have?

```{r}
penguins %>%
  count(species)
```

Here, using count(), we got the number of each species of penguins. There are 33 Gentoo penguins, 9 Adelie and 2 Chinstraps.

### Let's find the number of male and female penguins.

How many of each sex are there? What about by island or species?

```{r}
#number of male penguins
penguins %>%
  count(sex)

penguins %>%
  group_by(species) %>%
  count(sex)


```

Using #count(sex), we got the total numbers of female and male penguins. The number of males are higher than females. Also, using #group_by(species), we found out the number of males and females in each of the species.

We can use a function called `mutate()` to add new columns to our data set.

```{r}
penguins_with_ratio <- penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

#Average Ratio
penguins %>%
   mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
   summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio), 
          median_bill_ltd_ratio = median(bill_ltd_ratio))

#Average Ratio by Group
penguins %>%
  group_by(species) %>%
   mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
   summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio), 
          median_bill_ltd_ratio = median(bill_ltd_ratio))


```

Here, we got the average ratio of bill length to the bill depth. We used #mutate to make a different column for the data set. We got a mean and a median of the average bill length to bill depth ratio for all the species and each species as well respectively.

### Average body mass by year

```{r}
penguins %>%
  group_by(year) %>%
  summarize(mean_body_mass = mean(body_mass_g))
```

Using #summarize, we got the average body mass for 2007, 2008 and 2009. 2007 had the highest average body mass.

## Data Visualization

-   What is the distribution of penguin flipper lengths?

-   What is the distributive of penguin species?

-   Does the distribution of flipper length depend on the species of penguin?

    ```{r}
    penguins %>%
      ggplot() +
      geom_histogram( aes(x = flipper_length_mm),
                      bins = 15,
                      fill = "tan",
                      color = "Black") +
      labs(title = "Distribution of Flipper Length(mm)",
           y = "", x = "Flipper Length (mm)",
    subtitle = "Mean in Coral, Median in Blue") +
      geom_vline(aes(xintercept = mean
                 (flipper_length_mm)), lwd = 2, lty = "dashed", color = "coral"
      ) + 
                  geom_vline(aes(xintercept = median
                 (flipper_length_mm)), lwd = 2, lty = "dotted", 
                  color ="blue")

    ```

## We will now look at the distribution of species.

```{r}
penguins%>%
  ggplot()+
  geom_bar(mapping = aes(x = species), color ="black", fill = "lightblue" )+
  labs(title = "Counts of Penguin Species",
      x="Species", y = "Count")
```

Using #geom_bar, we constructed a bar graph to show the numbers of the species of the penguins. Gentoo has the highest number of species in this data set.

## Let's make a scatter plot to see if bill length is correlated with bill depth.

```{r}
penguins%>%
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm, color= species)) +
  geom_smooth(aes(x = bill_length_mm, y = bill_depth_mm, color= species), method = "lm")
```

Here, we made a scatterplot of the bill length and bill depth of each of the penguins. There is not enough data on the Chinstrap penguins so there is not much scatter points on the graph. The scatterplot showed a correlation between the bill lenght and the bill depth. They seem to be directly proportional.

### A Final Question

### We will find if the average bill length of a penguin exceeds 45mm or not.

```{r}
penguins%>%
  summarize(avg_bill_length = mean(bill_length_mm))
t.test(penguins$bill_length_mm, alternative= "greater" , mu= 45, conf.level= 0.95)
```

First we found out the average bill length. Then we had a hypothesis of whether the bill length of a penguin exceeded 45mm or not. Using confident interval and a T-test, the alternative hypothesis came out to be true. The null hypothesis was rejected. The average bill length of a penguin did exceed 45mm.

## Conclusion

In conclusion, the Palmer Penguin dataset provides valuable insights into the characteristics and distribution of different penguin species in the southern oceans. Through the analysis of various physical and ecological variables, researchers have been able to identify key differences between the Adelie, Chinstrap, and Gentoo penguin species. These differences can be used to inform conservation efforts and to better understand the impact of environmental changes on these species. The dataset also serves as a valuable resource for further studies in penguin ecology and evolution. Overall, the Palmer Penguin dataset represents a significant contribution to the field of marine biology and highlights the importance of interdisciplinary research in understanding and conserving the biodiversity of our planet.
